---
title: "2. Analyses"
author: "Lauke"
date: "28/01/2022"
output: html_document
---


```{r load in data}
load("~/GitHub/Master_Thesis/Analyses/Workspaces/step1_dat.Rdata")
```


```{r Libraries}
library(dplyr)
library(MCMCpack) #needed for PSM procedure MCMClogit function
library(tableone) #for CreateTableOne function
```

Step 1: obtain a propensity score distribution for each pupil. 

Variables to estimate PS on. NB: must be variables that were measured before treatment assignment (Kaplan and Chen 2012)
- ESCS
- age
- grade: captures whether the student is in the country’s modal grade (value of 0), or the number of grade below or above the modal grade in the country.
- ISCED level
- if the language of the test is the same language spoken at home. 

# Balance check 'before' - only dichotomy = 12
```{r}
table1 <-
  CreateTableOne(
    vars = c(
      "Gender",  
      "AGE",
      "GRADE", 
      "ISCEDL", #ISCED level 
      "ESCS", 
      "Language" #language at home vs. test
      ),
strata = "group12",
data = step1_dat,
test = FALSE
  )

print(table1, smd = T)
```
# PROPSENSITY SCORE ESTIMATION

## AGE 11
```{r Estimate posterior coefficients for alpha & betas - AGE 11}
set.seed(123)
posterior_11 <- MCMClogit(group11 ~ as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language), mcmc = 2000, burnin = 1000, thin = 2, data = step1_dat)

plot(posterior_11) #convergence is fine
summary(posterior_11)
```

```{r Estimate propensity scores for each pupil based on posterior samples of alpha and beta - AGE 11}
#create matrix containing all independent variables + intercept (1)
X_matrix_11 <- step1_dat %>% 
  mutate(constant = 1) %>% #add column to be multiplied by intercept
  dplyr::select(constant, Gender, AGE, GRADE, ISCEDL, ESCS, Language) %>% #select intercept + Xvars
  t() %>% #transpose to fit posterior object (col = pupils, row = vars)
  as.matrix() #store as matrix

#multiply the obtained posterior estimates with IV's to obtain propensity scores
prop_scores_11 <- exp(posterior_11 %*% X_matrix_11) / #logistic regression, see manuscript equation 3.
  (1 + exp(posterior_11 %*% X_matrix_11)) #col = pupils, row = n.iter prop scores

colnames(prop_scores_11) <- step1_dat$CNTSTUID #give cols student ID names

hist(colMeans(prop_scores_11))
mean(colMeans(prop_scores_11))
table(step1_dat$group11)
```

## AGE 12
```{r Estimate posterior coefficients for alpha & betas - AGE 12}
set.seed(123)
posterior_12 <- MCMClogit(group12 ~ as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language), mcmc = 2000, burnin = 1000, thin = 2, data = step1_dat)

plot(posterior_12) #convergence is fine
summary(posterior_12)
```

```{r Estimate propensity scores for each pupil based on posterior samples of alpha and beta - AGE 12}
#create matrix containing all independent variables + intercept (1)
X_matrix_12 <- step1_dat %>% 
  mutate(constant = 1) %>% #add column to be multiplied by intercept
  dplyr::select(constant, Gender, AGE, GRADE, ISCEDL, ESCS, Language) %>% #select intercept + Xvars
  t() %>% #transpose to fit posterior object (col = pupils, row = vars)
  as.matrix() #store as matrix

#multiply the obtained posterior estimates with IV's to obtain propensity scores
prop_scores_12 <- exp(posterior_12 %*% X_matrix_12) / #logistic regression, see manuscript equation 3.
  (1 + exp(posterior_12 %*% X_matrix_12)) #col = pupils, row = n.iter prop scores

colnames(prop_scores_12) <- step1_dat$CNTSTUID #give cols student ID names

hist(colMeans(prop_scores_12)) 
mean(colMeans(prop_scores_12)) 
table(step1_dat$group12)
```

## AGE 14
```{r Estimate posterior coefficients for alpha & betas - AGE 14}
set.seed(123)
posterior_14 <- MCMClogit(group14 ~ as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language), mcmc = 2000, burnin = 1000, thin = 2, data = step1_dat)

plot(posterior_14) #convergence is fine
summary(posterior_14)
```

```{r Estimate propensity scores for each pupil based on posterior samples of alpha and beta - AGE 14}
#create matrix containing all independent variables + intercept (1)
X_matrix_14 <- step1_dat %>% 
  mutate(constant = 1) %>% #add column to be multiplied by intercept
  dplyr::select(constant, Gender, AGE, GRADE, ISCEDL, ESCS, Language) %>% #select intercept + Xvars
  t() %>% #transpose to fit posterior object (col = pupils, row = vars)
  as.matrix() #store as matrix

#multiply the obtained posterior estimates with IV's to obtain propensity scores
prop_scores_14 <- exp(posterior_14 %*% X_matrix_14) / #logistic regression, see manuscript equation 3.
  (1 + exp(posterior_14 %*% X_matrix_14)) #col = pupils, row = n.iter prop scores

colnames(prop_scores_14) <- step1_dat$CNTSTUID #give cols student ID names

hist(colMeans(prop_scores_14)) 
mean(colMeans(prop_scores_14)) 
table(step1_dat$group14)
```

## AGE 15
```{r Estimate posterior coefficients for alpha & betas - AGE 15}
set.seed(123)
posterior_15 <- MCMClogit(group15 ~ as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language), mcmc = 2000, burnin = 1000, thin = 2, data = step1_dat)

plot(posterior_15) #convergence is fine
summary(posterior_15)
```

```{r Estimate propensity scores for each pupil based on posterior samples of alpha and beta - AGE 15}
#create matrix containing all independent variables + intercept (1)
X_matrix_15 <- step1_dat %>% 
  mutate(constant = 1) %>% #add column to be multiplied by intercept
  dplyr::select(constant, Gender, AGE, GRADE, ISCEDL, ESCS, Language) %>% #select intercept + Xvars
  t() %>% #transpose to fit posterior object (col = pupils, row = vars)
  as.matrix() #store as matrix

#multiply the obtained posterior estimates with IV's to obtain propensity scores
prop_scores_15 <- exp(posterior_15 %*% X_matrix_15) / #logistic regression, see manuscript equation 3.
  (1 + exp(posterior_15 %*% X_matrix_15)) #col = pupils, row = n.iter prop scores

colnames(prop_scores_15) <- step1_dat$CNTSTUID #give cols student ID names

hist(colMeans(prop_scores_15)) 
mean(colMeans(prop_scores_15)) 
table(step1_dat$group15)
```
```{r Save propensity score files}
save(prop_scores_11, file = "Workspaces/prop_scores_11.Rdata")
save(prop_scores_12, file = "Workspaces/prop_scores_12.Rdata")
save(prop_scores_14, file = "Workspaces/prop_scores_14.Rdata")
save(prop_scores_15, file = "Workspaces/prop_scores_15.Rdata")
```


# VISUALISATION (only for dichotomoy = 12)
```{r Check which students with prop scores belong to which group}
prop_scores_by_group <- data.frame(mean_propscores = colMeans(prop_scores_12), #mean ps per student
                         CNTSTUID = as.numeric(colnames(prop_scores_12))) %>%  #extract student ID
  left_join(step1_dat %>% dplyr::select(CNTSTUID, group12)) #join ps and student ID with group membership

# Create separate data files for T = 1 vs T = 0 
contr <- prop_scores_by_group %>% filter(group12 == 0)
treat <- prop_scores_by_group %>% filter(group12 == 1)
mean(contr$mean_propscores)
mean(treat$mean_propscores)


# Create histograms, then plot one and add the other: 
hist0 <- hist(contr$mean_propscores, breaks=30, plot=FALSE)
hist1 <- hist(treat$mean_propscores, breaks=30, plot=FALSE)
plot(hist1, col=rgb(0,0,1,1/4), xlim=c(0,1), 
     xlab="Propensity score",
     main="Histogram of propensity scores")  
plot(hist0, col=rgb(1,0,0,1/4), xlim=c(0,1), add=T) 

```

```{r Remove redudant objects from workspace}
rm(posterior, X_matrix, table1)
```





