---
title: "2.2 Outcome Model"
author: "Lauke"
date: "05/04/2022"
output: html_document
---

2. Obtain an estimate for the treatment effect

```{r}
set.seed(123)
```


```{r Load data}
load("~/GitHub/Master_Thesis/Analyses/Workspaces/fullmatchdat_p12_all.Rdata")
load("~/GitHub/Master_Thesis/Analyses/Workspaces/PVs.Rdata")
```


```{r Libraries}
library(haven)
library(dplyr)
library(MCMCpack)
library(mcmcplots)
library(knitr)
library(kableExtra)
```

Now we have 1000 different sets of matched units to pour into the analyses that estimates the treatment effect. First, we have to define the Y-variable on which we're regressing: ability.

# WITH FULLMATCH

```{r Create step2 matched data sets AGE 11}
#add ability estimate to each matched dataset for step 2
step2_dat_full_p11 <- lapply(fullmatchdat_p11_all, function(x) {inner_join(x, PVs[,c(1,12)], by = "CNTSTUID")})

save(step2_dat_full_p11, file = "Workspaces/step2dat_full_p11.Rdata")
str(step2_dat_full_p11[[1]]$subclass)
```

```{r Create step2 matched data sets AGE 12}
#add ability estimate to each matched dataset for step 2
step2_dat_full_p12 <- lapply(fullmatchdat_p12_all, function(x) {inner_join(x, PVs[,c(1,12)], by = "CNTSTUID")})

save(step2_dat_full_p12, file = "Workspaces/step2dat_full_p12.Rdata")
```

### TEST ENVIRONMENT
```{r test based on code by Chen on fullmatch object age 11}
test <- MCMCregress(
      ability ~ as.factor(group11) + as.factor(fullmatches_p11[[1]]$obj),
      mcmc = 50,
      burnin = 25, 
      verbose = 1, 
      data = step2_dat_full_p11[[1]]
    )[,2]
```

Chen code gave: Error in base::try(posterior_treat_full[[i]], silent = TRUE) :  object 'posterior_treat_full' not found.
Odd, stayed unsolved. When verbose = 1, it did start running, just very slow

 
```{r test on my solution to just as.factor(subclass)}
test2 <- MCMCregress(
      ability ~ as.factor(group11) + as.factor(subclass),
      mcmc = 20,
      burnin = 10, 
      verbose = 1,
      data = step2_dat_full_p11[[1]]
    )[,2]

plot(test2)
```
Again: Error in base::try(posterior_treat_full[[i]], silent = TRUE) :  object 'posterior_treat_full' not found. ?!?!?!?!

When posterior_treat_full was just removed from the picture entirely, it gave error: no loop for break/next, jumping to top level
with verbose = 1 it does run, just takes long

### REAL DEAL

DECISION: 
- I'll have 500 prop_score 12 matched datasets tonight
- I'll still present the average balance statisitcis of the whole thing.
- of every 50th matched dataset (so 10 in total), I estimate the treatment effect for 100 iterations (i.e. 1000 estimations in total)
- that's IT, done, donezo byebye. For ext 1 simply say, do this 4 times 

```{r Test on 1 dataset}
posterior_treat_full_p12_test <- lapply(step2_dat_full_p12[1], function(x) {
    MCMCregress(
      ability ~ as.factor(group11) + as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language) + as.factor(subclass),
      mcmc = 2,
      burnin = 0,
      verbose = 1,
      data = x
    )[,1:8]
  })

#2.6 minutes for 2 iterations on 1 dataset. but lot of starting time 
2.6*50 #130 minutes for 100 iterations
(130*10)/60 #21 hours for 100 iterations on 10 datasets??
```


```{r Bayesian regression estimation AGE 12}
set.seed(123)

detach("package:doSNOW", unload = T) #b/c it masks a shit tonne
library(parallel)

gc() #clean out unused memory
nclust <- parallel::detectCores() - 1
cl <- makeCluster(nclust)

#export objects to newly created R seshs
clusterExport(cl, c('step2_dat_full_p12'))
clusterEvalQ(cl, {(library("MCMCpack"))})

start_time <- Sys.time()

posterior_treat_full_p12 <- parLapply(cl, step2_dat_full_p12[c(seq(50,500, by = 50))], function(x) {
    MCMCregress(
      ability ~ as.factor(group12) + as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language) + as.factor(subclass),
      mcmc = 100,
      burnin = 0,
      verbose = 1,
      data = x
    )[,1:8] #extract only the first 8 regression coefficient per set
  })

end_time <- Sys.time()

stopCluster(cl)

comptime_posterior_12 <- end_time - start_time

save(posterior_treat_full_p12, file = "Workspaces/posterior_treat_full_p12.Rdata")
save(comptime_posterior_12, file = "Workspaces/comptime_posterior_p12.Rdata")
```

```{r Backup}
posterior_treat_full_p12 <- lapply(step2_dat_full_p12[c(seq(50,500, by = 50))], function(x) {
    MCMCregress(
      ability ~ as.factor(group12) + as.factor(Gender) + AGE + GRADE + ISCEDL + ESCS + as.factor(Language) + as.factor(subclass),
      mcmc = 100,
      burnin = 0,
      verbose = 1,
      data = x
    )[,1:8]
  })
```


```{r Quick check on convergence, echo = F}
posterior_treat_full_p12 %>% lapply(., traplot)
posterior_treat_full_p12 %>% lapply(., autocorr.plot)
```

```{r Final estimates - regular}
summary(posterior_treat_full_p12[[1]]) #treatment effect -16.344, significant

#average posterior mean:
posterior_treat_full_p12 %>% lapply(., colMeans) %>% do.call(rbind, .) %>% as.data.frame() %>% colMeans

#CCI WRONG - OLD (first pull together then quantile whole thing)
#posterior_treat_full_p12 %>% do.call(rbind, .) %>% as.data.frame() %>% apply(., 2, quantile, 0.025)
#posterior_treat_full_p12 %>% do.call(rbind, .) %>% as.data.frame() %>% apply(., 2, quantile, 0.975)

#CCI: CORRECT - first quantile each then average 
posterior_treat_full_p12 %>% lapply(., function(x) {apply(x, 2, quantile, 0.025)}) %>% do.call(rbind, .) %>% colMeans
posterior_treat_full_p12 %>% lapply(., function(x) {apply(x, 2, quantile, 0.975)}) %>% do.call(rbind, .) %>% colMeans

#sd: 
posterior_treat_full_p12 %>% do.call(rbind, .) %>% as.data.frame() %>% apply(., 2, sd)


results <- matrix(0, 8, 4)
rownames(results) <- c("Intercept", "Treatment", "Gender", "AGE", "GRADE", "ISCEDL", "ESCS", "Language")
colnames(results) <- c("Average Posterior Mean", "2.5%",  "97.5%", "Average Posterior SD")


results[, 1] <- posterior_treat_full_p12 %>% lapply(., colMeans) %>% do.call(rbind, .) %>% as.data.frame() %>% colMeans
results[, 2] <- posterior_treat_full_p12 %>% lapply(., function(x) {apply(x, 2, quantile, 0.025)}) %>% do.call(rbind, .) %>% colMeans
results[, 3] <- posterior_treat_full_p12 %>% lapply(., function(x) {apply(x, 2, quantile, 0.975)}) %>% do.call(rbind, .) %>% colMeans
results[, 4] <- posterior_treat_full_p12 %>% do.call(rbind, .) %>% as.data.frame() %>% apply(., 2, sd)

kable(results, digits = 2)

kbl(as.data.frame(results), 
    booktabs = T, 
    digits = 2, 
    format = "latex",
    caption = "Final estimates Bayesian regression", 
    label = "outcome_norm")
```

```{r}
hist(posterior_treat_full_p12[[1]][,2])
```





